---
title: "Supplementary information 1"
author: 
  - name: Anaïs L. Vignoles
    affiliation: University of Cologne
    orcid: 0000-0001-5027-2037
  - name: Shumon T. Hussain
    affiliation: University of Cologne
    orcid: 0000-0002-6215-393X
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    fig_caption: yes
    theme: sandstone
    highlight: zenburn
bibliography: "biblio-suppl.bib"
link_citations: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, fig.width = 10)

```

This supplementary document groups together data and code for the
analysis presented in Hussain et al. *Technospheric Dynamics in the Deep
Human Past: Multi-Level Causation in the Macroevolution of Late
Pleistocene Lithic Technology*. One can access code and data of this
supplementary material through [this
link](https://github.com/ALVignoles/Technological-holarchies) 

The packages used in this paper are listed below. 

```{r packages}

library(viridis)
library(nortest)
library(onewaytests)

# this helper function for replacing values based on a correspondence table is also coded here 
replace_values <- function(corresp.table, word1, word2, replaced.values){
  for(i in 1:nrow(corresp.table)) {
  search_keyword <- corresp.table[i,word1]
  replace_keyword <- corresp.table[i, word2]
  
  replaced.values <- sapply(replaced.values, function(x){
    x[x==search_keyword] <- replace_keyword
    return(x)
  })
  }
  return(replaced.values)
}

```

# Presentation of the data set

We re-analyze the data shared in the framework of a study conducted by
[Rezek et al. in 2018](https://doi.org/10.1038/s41559-018-0488-4). Two
csv tables and a pdf were downloaded at
<https://www.nature.com/articles/s41559-018-0488-4#Sec9>. Before we
present our re-analysis, we first summarize Rezek et al. 2018
methodology to build the data set and provide a critical opinion on the
quality of the data set. 

## Methodology

The data set consists of flake size measurements for 81 assemblages from
34 sites in Africa, south-west Asia and western Europe. Only unmodified
and complete flakes were taken into account; their width, length and
thickness were measured, as well as other traits that will not be
discussed here. Only flakes larger/longer than 25 mm were measured. Abri
Pataud assemblages have not been measured in totality, but only through
a random sample of at least 100 flakes per layer. No contextual
information is given, except: 

*"Many of the assemblages were excavated using modern archaeological
standards"* [@Rezek2018, p.1]. 

The proxy used in Rezek et al. to infer cognition evolution is the ratio
between sharp edge amount and flake size, hereby referred to as EFS
ratio. 

## A critique of Rezek et al. approach to macro-archaeology data sets building

Before re-using the data set provided in Rezek et al. 2018, we would
like to state a number of issues that, in our opinion, compromises the
quality, and therefore the reusability, of this data set. Because our
literature review was mostly focuses on European sites, the comments
will rely more on examples from South-Western France assemblages. 

- **Cultural attributions are sometimes erroneous**, and not in line
  with research published at the time of submission (18-09-2017). For
  example, Quina-type *débitage* is not demonstrated in level 3 at Roc
  de Marsal, where Levallois and Laminaire have been described
  [@Turq2017]. Moreover, level 31 at Laugerie Haute-Est is also not
  generally considered Lower Solutrean, but Proto-Solutrean
  [@Demars1995].\

- From a general point of view, **many of these assemblages have not
  been revised recently**, which makes it difficult to assess the
  quality of chrono-cultural attribution. Level 40 of Laugerie Haute-Est
  is a good example, its lithic industry has never been reviewed since
  its initial publication in 1958 by F. Bordes [@Cormareche2024, p.172].

- In addition, altough *"many of the assemblages were excavated using
  modern archaeological standards"* [@Rezek2018, p.1], **excavation
  methodology remains highly heterogeneous**, ranging from 1912 to 2014.
  Most assemblages were excavated in the 1950-1980s, a period
  characterized by an important variability in excavation practices that
  have been shown to not always correspond to modern standards [for a
  review targeting the Middle and Recent Gravettian, see @Vignoles2024].
  The example of Pataud is striking: Movius' excavations are often
  considered highly qualitative, and they were indeed visionary for the
  period during which they took place [@Movius1975]. However, recent
  excavations of layer 2 led by R. Nespoulet and L. Chiotti in the 2000s
  have shown that the Movius assemblage lacks smaller items that are
  essential to interpret Final Gravettian llithic industries for example
  [@Nespoulet2008]. It is also well known that the Fourneau-du-Diable
  assemblage from the Peyrony collections is highly sorted, with little
  regard to unretouched material [@Baumann2016; @Vignoles2019].

- **Levels homogeneity is not discussed** while some levels are clearly
  mixed OR have not been assessed while we know that such contexts can
  be mixed (mainly rock shelters). Most assemblages (if not all) from
  Upper Paleolithic contexts are involved. For example, level 8 at Roc
  de Combe presents several intrusions from level 7 (Aurignacian) and
  level 9 (Mousterian) [@Bordes2003]. Another striking example is that
  of Fourneau du Diable Peyrony collection, which is a highly sorted and
  mixed collection [@Vignoles2019] that might contain a Magdalenian
  component [@Baumann2016]. This is also probably the case at other
  rockshelter sites that have not been revised from a contextual point
  of view, such as La Madeleine (where levels 5-6 were interpreted as
  interface layers; [@Bundgen2002]) or Laugerie Haute-Est.  

- **Some of these assemblages have been demonstrated to group different
  archaeological units together**. For example, the Middle Gravettian is
  divided into Noaillian and Rayssian which have been shown to have
  different typo-technical characteristics [@David1985; @Klaric2007].
  The case is also true for the Badegoulian that has an internal phasing
  [@Cretin2007]. Finally, some assemblages have been classified as
  different subdivisions of techno-complexes while they are highly
  similar from a typo-technical standpoint (layers 30 and 28-21 at
  Laugerie Haute-est, [@Demars1995]). 

- **The measurements were made on artefacts bigger than 2cm, which
  undermines greatly the variability of industries that have a strong
  bladelet component**, such as the Final Gravettian in particular
  [@Nespoulet2008]. 

- Finally, **Cited articles for a chronocultural attribution do not
  actually demonstrate these attributions** from the point of view of
  the material culture, but rather based on dating articles or old
  typological analysis (eg., level 5 of Abri Pataud; Roc de Combe layer
  8, for which articles referring to Mousterian and Aurignacian
  assemblages are cited [@Bordes2003; @Lorenzo2014]). 

# Exploring the EFS ratio variability according to chrono-cultural attributions of assemblages

In this section, we explore flake size variability according to
chrono-cultural attributions of assemblages. 

## Data preparation

We first load the two tables provided in the supplementary data and
correct a couple of typos in the tables. 

```{r load}

data <- read.csv("41559_2018_488_MOESM3_ESM.csv")
head(data)

rezek_summary <- read.csv("41559_2018_488_MOESM4_ESM.csv")
head(rezek_summary)

# correcting typos
rezek_summary$Assemblage.code[rezek_summary$Assemblage.code == "W-A-CC"] <- "W-A-CCC"
rezek_summary$Archaeological.industry[rezek_summary$Archaeological.industry == "Middle Gravetian"] <- "Middle Gravettian"

```

When exploring the two tables, we can see that chrono-cultural
attributions of individual artefacts in the "MOESM3" table and of
individual levels in the "MOESM4" table are different. Considering the
later to be more accurate (contains all of the subdivisions of the Upper
Palaoelithic for eg.), we re attribute individual artefacts of table
"MOESM3" to their level attribution in table "MOESM4". 

```{r homogeneize}

data$industry_MOESM4 <- replace_values(rezek_summary, "Assemblage.code", "Archaeological.industry", data$Assemblage.code)

head(data)

```

Data are then ordered according to chrono-cultural attributions... 

```{r order}

unique(data$industry_MOESM4)
industries <- c("Oldowan", "Oldowan (Developed)", "Acheulian", "Karari", "Khormusan", "Middle Stone Age of NW Africa",
                
                "Levantine Mousterian", "Zagros Mousterian", "Ahmarian", "Levantine Aurignacian",                                    
                "Mousterian", "Quina Mousterian", "Mousterian of Acheulian Tradition", "Chatelperronian", "Early Aurignacian", 
                "Recent Aurignacian", "Early Gravettian", "Middle Gravettian", "Recent Gravettian", "Final Gravettian" , 
                "Lower Solutrean", "Middle Solutrean", "Upper Solutrean", "Badegoulian", "Lower Magdalenian", 
                "Middle Magdalenian", "Upper Magdalenian")

```

... and a color palette is coded for the three regions. 

```{r pal1}

colors <- industries

colors[1:6] <- cividis(3)[3]
colors[7:10] <- cividis(3)[1]
colors[11:length(colors)] <- cividis(3)[2]

```

## Summarizing EFS ratio data for each archaeological unit

Since median EFS ratio for individual assemblages was the only available
value in table "MOESM4", we recalculate quantiles, as well as variance,
for each chrono-cultural attribution. 

```{r summ-arch}

summary_industry <- data.frame(industry = NA, 
                      ratio_min = NA, ratio_25 = NA, median = NA, ratio_75 = NA, ratio_max = NA, var = NA)

for(i in 1:length(industries)){
  assemb <- subset(data, data$industry_MOESM4 == industries[i])
  
  ratio <- assemb$length.width.thickness.2
  ratio <- na.omit(as.numeric(ratio))
  
  q <- quantile(ratio)
  
  var <- var(ratio)
  
  sum <- c(assemb$industry_MOESM4[1], q, var)
  summary_industry <- na.omit(rbind(summary_industry, sum))
}

summary_industry$industry <- factor(summary_industry$industry, levels = industries)

head(summary_industry)

```

The data is ready to be visualized. 

## Views of the data

Next are different views of the data ordered in different ways. 

### EFS ratio distributions ordered by chronology and region

```{r chronology}

summary_industry <- summary_industry[order(summary_industry$industry),]
summary_industry$col <- colors
summary_industry$y <- c(1:nrow(summary_industry))

plot(x = 1, col = "white", xlim = c(0,550), ylim = c(0,28),
     xlab = "sharp edge/flake size ratio", ylab = "", yaxt = "n", bty = "n")

segments(x0 = as.numeric(summary_industry$ratio_min), x1 = as.numeric(summary_industry$ratio_max), y0 = summary_industry$y, lwd = 1.5)
segments(x0 = as.numeric(summary_industry$ratio_25), x1 = as.numeric(summary_industry$ratio_75), y0 = summary_industry$y, lwd = 4)
points(x = summary_industry$median, y = summary_industry$y,
       pch = 23, cex=1.3, bg = summary_industry$col)

text(x=as.numeric(summary_industry$ratio_max)+10, y=summary_industry$y, labels = summary_industry$industry, adj = 0, 
     cex = 0.8)
legend(x = 420, y = 28, legend = c("Africa", "Near East", "Europe"), cex = 0.7, fill = c(cividis(3)[3], cividis(3)[c(1,2)]))

```

### EFS ratio distributions ordered by median value

```{r median}
summary_median <- summary_industry[order(as.numeric(summary_industry$median)),]
summary_median$y <- c(1:nrow(summary_median))

plot(x = 1, col = "white", xlim = c(0,550), ylim = c(0,28),
     xlab = "sharp edge/flake size ratio", ylab = "", yaxt = "n", bty = "n")

segments(x0 = as.numeric(summary_median$ratio_min), x1 = as.numeric(summary_median$ratio_max), y0 = summary_median$y, lwd = 1.5)
segments(x0 = as.numeric(summary_median$ratio_25), x1 = as.numeric(summary_median$ratio_75), y0 = summary_median$y, lwd = 4)
points(x = summary_median$median, y = summary_median$y,
       pch = 23, cex=1.3, bg = summary_median$col)

text(x=as.numeric(summary_median$ratio_max)+10, y=summary_median$y, labels = summary_median$industry, adj = 0, 
     cex = 0.8)
legend(x = 420, y = 8, legend = c("Africa", "Near East", "Europe"), cex = 0.7, fill = c(cividis(3)[3], cividis(3)[c(1,2)]))

```

### EFS ratio distributions ordered by ecartype (interquantile range)

```{r ecartype}
summary_industry$et <- as.numeric(summary_industry$ratio_75) - as.numeric(summary_industry$ratio_25)
summary_ecartype <- summary_industry[order(as.numeric(summary_industry$et)),]
summary_ecartype$y <- c(1:nrow(summary_ecartype))

plot(x = 1, col = "white", xlim = c(0,550), ylim = c(0,28),
     xlab = "sharp edge/flake size ratio", ylab = "", yaxt = 'n', bty = "n")

segments(x0 = as.numeric(summary_ecartype$ratio_min), x1 = as.numeric(summary_ecartype$ratio_max), y0 = summary_ecartype$y, lwd = 1.5)
segments(x0 = as.numeric(summary_ecartype$ratio_25), x1 = as.numeric(summary_ecartype$ratio_75), y0 = summary_ecartype$y, lwd = 4)
points(x = summary_ecartype$median, y = summary_ecartype$y,
       pch = 23, cex=1.3, bg = summary_ecartype$col)

text(x=as.numeric(summary_ecartype$ratio_max)+10, y=summary_ecartype$y, labels = summary_ecartype$industry, adj = 0, 
cex = 0.8)
legend(x = 420, y = 4.5, legend = c("Africa", "Near East", "Europe"), cex = 0.7, fill = c(cividis(3)[3], cividis(3)[c(1,2)]))

```

### EFS ratio distributions ordered by full range

```{r range}
summary_industry$rg <- as.numeric(summary_industry$ratio_max) - as.numeric(summary_industry$ratio_min)
summary_range <- summary_industry[order(as.numeric(summary_industry$rg)),]
summary_range$y <- c(1:nrow(summary_range))

plot(x = 1, col = "white", xlim = c(0,550), ylim = c(0,28),
     xlab = "sharp edge/flake size ratio", ylab = "", yaxt = "n", bty = "n")

segments(x0 = as.numeric(summary_range$ratio_min), x1 = as.numeric(summary_range$ratio_max), y0 = summary_range$y, lwd = 1.5)
segments(x0 = as.numeric(summary_range$ratio_25), x1 = as.numeric(summary_range$ratio_75), y0 = summary_range$y, lwd = 4)
points(x = summary_range$median, y = summary_range$y,
       pch = 23, cex=1.3, bg = summary_range$col)

text(x=as.numeric(summary_range$ratio_max)+10, y=summary_range$y, labels = summary_range$industry, adj = 0, 
     cex = 0.8)
legend(x = 420, y = 5, legend = c("Africa", "Near East", "Europe"), cex = 0.7, fill = c(cividis(3)[3], cividis(3)[c(1,2)]))
```

# A zoom on Europe: incorporating technology into the data set

In this section, we experiment incorporating technological information
in the data set, by associating technical systems identification to
individual assemblages. We conducted this work solely on the European
sites because of limited time devoted to analysis. All sites are located
in Northern Aquitaine region, which in addition gives a better
geographic coherence to the results. 

The data set is then filtered to keep European sites... 

```{r europe}

data_europe <- data[data$industry_MOESM4 %in% c("Mousterian", "Quina Mousterian", "Mousterian of Acheulian Tradition", "Chatelperronian", "Early Aurignacian", 
                                         "Recent Aurignacian", "Early Gravettian", "Middle Gravettian", "Recent Gravettian", "Final Gravettian" , 
                                         "Lower Solutrean", "Middle Solutrean", "Upper Solutrean", "Badegoulian", "Lower Magdalenian", 
                                         "Middle Magdalenian", "Upper Magdalenian"),]
```

... and assemblages are recoded to be associated with higher level
system (techno-complexes), to simplify some of the figures. 

```{r TCL-recode}
recode.ind <- data.frame(industry = c("Mousterian of Acheulian Tradition", "Early Aurignacian", "Recent Aurignacian", 
                                      "Early Gravettian", "Middle Gravettian", "Recent Gravettian", "Final Gravettian", 
                                      "Lower Solutrean", "Middle Solutrean", "Upper Solutrean",
                                      "Lower Magdalenian", "Middle Magdalenian", "Upper Magdalenian"),
                         TCL = c("MTA", "Aurignacian", "Aurignacian",
                                 "Gravettian", "Gravettian", "Gravettian", "Gravettian", 
                                 "Solutrean", "Solutrean", "Solutrean",
                                 "Magdalenian", "Magdalenian", "Magdalenian"))

data_europe$TCL <- replace_values(recode.ind, "industry", "TCL", as.character(data_europe$industry_MOESM4))

data_europe <- data_europe[!is.na(data_europe$length.width.thickness.2),] 
head(data_europe)

```

## A literature review of individual levels technological characteristics

We conducted a literature review of individual levels technological
characteristics. We started from references listed in the supplementary
material pdf file, as well as recent academic work and synthesis papers.
When technological data was unavailable, we estimated presence of
technical systems (*eg.*, miscellaneous blade and bladelet *débitages*)
from drawings and general typological descriptions in older body of
literature. We also provide some contextual information and precision
about research biases whenever we deemed necessary, in order to better
assess the quality of information gathered in the literature. The result
of this work is available in Supplementary 2. 

## Coding technological data resulting from literature review

Following Supplementary 2, we code technological information in
the following way:

### Full data set

For each assemblage, we code information for 4 types of technological
systems: bifacial shaping, flake systems, blade systems and bledelet
systems. The data was recorded as pres, abs and na (not available). 

```{r code-techno}

assemb <- c("AP-U2", "AP-L3B", "AP-U3A", "AP-L3A", "RDM5", "RDM6", "RDM7", "RDM8", "RDM9",
            "PDA4C", "PDA5A", "PDA5B", "PDA6A", "PDA6B", "PDA8", "RDM2", "RDM3", "RDM4",
            "PDA4A", "CCI1D", "CCI1E", "CCI1D1", "CCI2A", "CCI2B", "CCII4A", "CCII4B", "CCII4C",  
            "CCII4E", "PDA3A", "PDA3B", "COT3", "RC8", "AP14", "AP13", "AP12", "AP11", "AP9-10",  
            "AP8", "AP7", "AP6", "AP5", "AP4", "AP3", "LHE40", "AP2", "LHE36-38", "LHE31", "LHE30", 
            "LHE21-28", "FD-U", "LHE18-20", "LHE10-16", "LHE4-8", "LHE2-3", "MAD5-6")

techno.data <- data.frame(assemb = assemb, 
                          bifacial = c("Bif", "Bif", "Bif","Bif", # Abri Peyrony
                                       "abs", "abs", "abs", "abs", "Bif", #  Roc de Marsal
                                       "abs", "abs", "abs", "abs", "abs", "abs", # Pech de l'Aze IV
                                       "abs", "abs", "abs", # Roc de Marsal
                                       "abs", # Pech de l'Aze IV
                                       "abs", "abs", "abs", "abs", "abs", "abs", "abs", "abs", "abs", # Combe Capelle bas
                                       "Bif", "Bif", # Pech de l'Aze IV
                                       "abs", # La Cote
                                       "abs", # Roc de Combe
                                       "abs", "abs", "abs", "abs", "abs", # Abri Pataud 14-9
                                       "abs", "abs", "abs", # Abri Pataud 6-8 
                                       "abs", # Abri Pataud 5
                                       "abs", # Abri Pataud 4
                                       "abs", # Abri Pataud 3
                                       "abs", # Laugerie Haute E
                                       "abs", # Abri Pataud 2 
                                       "abs", # Laugerie Haute E
                                       "na", "na", "na", # Laugerie Haute E
                                       "solut bif", # FDD
                                       "na", "na",
                                       "na", "na", "na"
                          ),
                          eclats = c("na", "na", "na", "na", # Abri Peyrony
                                     "Levallois", "Levallois", "Levallois", "Quina", "Quina", #  Roc de Marsal
                                     "Levallois, Kombewa", "Levallois, Kombewa", "Levallois, Kombewa", "Levallois, Kombewa", "Levallois, Kombewa", "Levallois, Kombewa", # Pech de l'Aze IV
                                     "Levallois", "Levallois", "Levallois", # Roc de Marsal
                                     "na", # Pech de l'Aze IV
                                     "Quina", "Quina", "Quina", "Quina", "Quina", "Quina", "Quina", "Quina", "Quina", # Combe Capelle Bas
                                     "Levallois, Discoid", "Levallois, Discoid", # Pech de l'Aze IV
                                     "abs", # La Cote
                                     "abs", # Roc de Combe
                                     "eclats", "eclats", "eclats", "eclats", "eclats", # Abri Pataud 14-9
                                     "eclats", "eclats", "eclats", # Abri Pataud 6-8
                                     "na", # Abri Pataud 5
                                     "eclats", # Abri Pataud 4
                                     "aneg", #  Abri Pataud 3
                                     "na", # Laugerie Haute E
                                     "aneg", # Abri Pataud 2
                                     "na", # Laugerie Haute E
                                     "na", "na", "na", # Laugerie Haute E
                                     "na", # FDD
                                     "debitage d'eclats normalises, eclats laminaires", "debitage d'eclats normalises, eclats laminaires", 
                                     "na", "na", "na"
                          ),
                          laminaire = c("abs", "abs", "abs", "abs", # Abri Peyrony
                                        "abs", "abs", "abs", "abs", "abs", #  Roc de Marsal
                                        "abs", "abs", "abs", "abs", "abs", "MP Lamn", # Pech de l'Aze IV
                                        "abs", "MP Lamn", "abs", # Roc de Marsal
                                        "abs", # Pech de l'Aze IV
                                        "abs", "abs", "abs", "abs", "abs", "abs", "abs", "abs", "abs", # Combe Capelle bas
                                        "abs", "abs", # Pech de l'Aze IV
                                        "Frontal, semi-tournant, en amande", # La Cote
                                        "Frontal, semi-tournant, en amande", # Roc de Combe
                                        "prismatique", "prismatique", "prismatique", "prismatique", "prismatique", # Abri Pataud 14-9
                                        "unipolaire", "unipolaire", "unipolaire", # Abri Pataud 6-8
                                        "pres", # Abri Pataud 5
                                        "unipolaire", # Abri Pataud 4
                                        "frontal uni bipolaire", # Abri Pataud 3
                                        "pres", # Laugerie Haute E
                                        "grandes lames eperon", # Abri Pataud 2
                                        "pres", # Laugerie Haute E
                                        "pres", "pres", "pres", # Laugerie Haute E
                                        "pres", # FDD
                                        "pres", "pres",
                                        "pres", "pres", "na"
                          ),
                          lamellaire = c("abs", "abs", "abs", "abs", # Abri Peyrony
                                         "abs", "abs", "abs", "abs", "abs", #  Roc de Marsal
                                         "abs", "abs", "abs", "abs", "abs", "abs", # Pech de l'Aze IV
                                         "abs", "abs", "abs", # Roc de Marsal
                                         "abs", # Pech de l'Aze IV
                                         "abs", "abs", "abs", "abs", "abs", "abs", "abs", "abs", "abs", # Combe Capelle bas
                                         "abs", "abs", # Pech de l'Aze IV
                                         "tranche d'eclat", # La Cote
                                         "na", # Roc de Combe
                                         "carene, prismatique, intercale", "carene, prismatique, intercale", "carene, prismatique, intercale", "carene, prismatique, intercale", "carene, prismatique, intercale", # Abri Pataud 14-9
                                         "carene", "carene, busque, vachons", "carene, busque, vachons", # Abri Pataud 6-8
                                         "pres", # Abri Pataud 5
                                         "methode Raysse, burins-nucleus", # Abri Pataud 4
                                         "continuum, burins-nucleus", # Abri Pataud 3
                                         "pres", # Laugerie Haute E
                                         "lamel", # Abri Pataud 2
                                         "pres", # Laugerie Haute E
                                         "pres", "pres", "pres", # Laugerie Haute E
                                         "pres", # FDD
                                         "burins tranversaux", "burins tranversaux",
                                         "pres", "pres", "na"
                          )
                          )
                          
head(techno.data)

```

### Summary data set

Because of the uneveness in the precision of technological information
gathered in the literature, we simplify these information at a higher
level, by considering main families of systems: bifacial shaping,
Levallois, Quina, Kombewa, Discoid, other flake *débitages*, Blade
*débitages*, Bladelet *débitages*. We obtain a summary of main families
of systems for each assemblage. 

```{r summary-techno}

techno.data$summary <- c("Bif", "Bif", "Bif","Bif", # Abri Peyrony
                         "Lev", "Lev", "Lev", "Q", "Bif+Q", #  Roc de Marsal
                         "Lev+Komb", "Lev+Komb", "Lev+Komb", "Lev+Komb", "Lev+Komb", "Lev+Komb+Blade", # Pech de l'Aze IV
                         "Lev", "Lev+Blade", "Lev", # Roc de Marsal
                         "", # Pech de l'Aze IV
                         "Q", "Q", "Q", "Q", "Q", "Q", "Q", "Q", "Q", # Combe Capelle bas
                         "Bif+Lev+D", "Bif+Lev+D", # Pech de l'Aze IV
                         "Blade+Bladelet", # La Cote
                         "Blade+Bladelet", # Roc de Combe
                         "Flake+Blade+Bladelet", "Flake+Blade+Bladelet", "Flake+Blade+Bladelet", "Flake+Blade+Bladelet", "Flake+Blade+Bladelet", # Abri Pataud 14-9
                         "Flake+Blade+Bladelet", "Flake+Blade+Bladelet", "Flake+Blade+Bladelet", # Abri Pataud 6-8 
                         "Blade+Bladelet", # Abri Pataud 5
                         "Flake+Blade+Bladelet", # Abri Pataud 4
                         "Blade+Bladelet", # Abri Pataud 3
                         "Blade+Bladelet", # Laugerie Haute E
                         "Blade+Bladelet", # Abri Pataud 2 
                         "Blade+Bladelet", # Laugerie Haute E
                         "Blade+Bladelet", "Blade+Bladelet", "Bif+Blade+Bladelet", # Laugerie Haute E
                         "Bif+Blade+Bladelet", # FDD
                         "Flake+Blade", "Flake+Blade", # Badegoulian 
                         "Blade+Bladelet", "Blade+Bladelet", "") # Magdalenian

unique(techno.data$summary)

```

### Recoding data for each assemblage

We then recode individual flake measurements to associate each one of
them to technological systems identification. 

```{r recode}

data_europe$bifacial <- replace_values(techno.data, "assemb", "bifacial", data_europe$Assemblage.code)

data_europe$eclats <- replace_values(techno.data, "assemb", "eclats", data_europe$Assemblage.code)

data_europe$laminaire <- replace_values(techno.data, "assemb", "laminaire", data_europe$Assemblage.code)

data_europe$lamellaire <- replace_values(techno.data, "assemb", "lamellaire", data_europe$Assemblage.code)

data_europe$summary <- replace_values(techno.data, "assemb", "summary", data_europe$Assemblage.code)

head(data_europe)

```

## Summarizing ratio data for each assemblage

We summarize EFS ratio for each assemblage. 

```{r ratio-techn}

summary_assemb <- data.frame(assemb = NA, bifacial = NA, eclats = NA, laminaire = NA, lamellaire = NA,
                             techn_syst = NA, industry = NA, TCL = NA,
                             ratio_min = NA, ratio_25 = NA, median = NA, ratio_75 = NA, ratio_max = NA, var = NA)

for(i in 1:length(assemb)){
  df <- subset(data_europe, data_europe$Assemblage.code == assemb[i])
  
  ratio <- df$length.width.thickness.2
  ratio <- na.omit(as.numeric(ratio))
  
  q <- quantile(ratio)
  
  var <- var(ratio)
  
  sum <- c(df$Assemblage.code[1], df$bifacial[1], df$eclats[1], df$laminaire[1], df$lamellaire[1], 
           df$summary[1], paste(df$industry_MOESM4[1]), df$TCL[1], q, var)
  summary_assemb <- na.omit(rbind(summary_assemb, sum))
}

summary_assemb$assemb <- factor(summary_assemb$assemb, levels = assemb)
summary_assemb <- summary_assemb[order(summary_assemb$techn_syst),]
summary_assemb$y <- c(1:nrow(summary_assemb))
```

We also code colors associated with each technical system, and shapes
associated with each techno-complex. 

```{r pal2}

colors <- turbo(length(unique(summary_assemb$techn_syst)))
summary_assemb$colors <- summary_assemb$techn_syst

for(i in 1:length(colors)) {
  search_keyword <- unique(summary_assemb$techn_syst)[i]
  replace_keyword <- colors[i]
  
  summary_assemb$colors <- sapply(summary_assemb$colors, function(x){
    x[x==search_keyword] <- replace_keyword
    return(x)
  })
}

shapes <- c(5, 3, 5, 4, 5, 0, 2, 6, 1)
summary_assemb$shapes <- summary_assemb$TCL

for(i in 1:length(shapes)) {
  search_keyword <- unique(summary_assemb$TCL)[i]
  replace_keyword <- shapes[i]
  
  summary_assemb$shapes <- sapply(summary_assemb$shapes, function(x){
    x[x==search_keyword] <- replace_keyword
    return(x)
  })
}

```

The data is ready to be visualized.  

## Views of the data

Next are different views of the data. 

### EFS ratio distributions for individual assemblages, ordered by technical system identification

```{r techn-syst, fig.height=6}

plot(x = 1, col = "white", xlim = c(0,550), ylim = c(0,57),
     xlab = "sharp edge/flake size ratio", ylab = "", yaxt = "n", bty = "n")

segments(x0 = as.numeric(summary_assemb$ratio_min), x1 = as.numeric(summary_assemb$ratio_max), y0 = summary_assemb$y, lwd = 1.5)
segments(x0 = as.numeric(summary_assemb$ratio_25), x1 = as.numeric(summary_assemb$ratio_75), y0 = summary_assemb$y, lwd = 4)
points(x = summary_assemb$median, y = summary_assemb$y,
       pch = 23, cex=1.3, bg = summary_assemb$colors)

text(x=as.numeric(summary_assemb$ratio_max)+10, y=summary_assemb$y, labels = summary_assemb$assemb, adj = 0, 
     cex = 0.8)
legend(x = 400, y = 50, legend = rev(unique(summary_assemb$techn_syst)), cex = 0.7, fill = rev(colors))


```

### Median and ecartype of EFS ratio distributions for individual assemblages

```{r med-rang, fig.height=6}
par(mfrow = c(1,1), mar = c(5.1, 4.1, 4.1, 2.1) )
    
summary_assemb$et_rg <- as.numeric(summary_assemb$ratio_75) - as.numeric(summary_assemb$ratio_25)

plot(x = 1, col = "white", xlim = c(0,100), ylim = c(0,57),
     xlab = "sharp edge/flake size ratio", ylab = "", yaxt = "n", bty = "n")


segments(x0 = as.numeric(summary_assemb$ratio_25), x1 = as.numeric(summary_assemb$ratio_75), y0 = summary_assemb$y, lwd = 1)
points(x = summary_assemb$median, y = summary_assemb$y,
       pch = as.numeric(summary_assemb$shapes), cex=1.3, lwd = 3,
       col = summary_assemb$colors)

legend(x = 71.5, y = 35, legend = rev(unique(summary_assemb$techn_syst)), cex = 0.8, fill = rev(colors))
legend(x = 71.5, y = 55, legend = c("Mousterian", "Chatelperronian", "Aurignacian", "Gravettian",
                                   "Solutrean", "Badegoulian", "Magdalenian"), cex = 0.8, 
       pch = c(5, 0, 1, 2, 4, 6, 3), col = "black")

```

## Tests for statistical difference

We analize these data further to see whether groups based on
technological systems identification and techno-complex have
significantly different distributions of EFS ratio. Since the
technological data and techno-complex attribution are independent from
individual flake measurements, we can explore whether these groups make
sense from a point of view of the EFS ratio distributions. 

### Data grouped by technological category

We first explore morphometric coherence of technological categories. 

Before running statistical differentiation tests, we check for each
groups' distribution normality via a Lilliefors test [@Lilliefors1967]. 

```{r normal1, fig.height=10}

par(mfrow = c(4,3))

for(i in 1:length(unique(techno.data$summary))) {
  groupn <- data_europe[data_europe$summary == unique(techno.data$summary)[i], "length.width.thickness.2"]
  
  title <- paste(unique(techno.data$summary)[i])
  
  #Lilliefors test
  pvalue <- signif(lillie.test(groupn)$p.value, digits = 3)
  sample.size <- length(groupn)
  
  hist(groupn, breaks = seq(0,500,10), main = title, xlab = "sharp edge/flake size ratio", col = cividis(3)[2])
  legend(x = .25 * par('usr')[2], y = .9 * par('usr')[4], 
         legend = c(paste0("Lilliefors test p-value: ", pvalue), paste0("sample size: ", sample.size)), bty = "n")
}


```

All groups present p-values under 0.003 [@Courtenay2024] except for the
Levallois + Blade group. It is therefore removed from further analysis,
as well as assemblages with no technological information available. 

```{r remove}

data_europe_red <- data_europe[data_europe$summary != "",]
data_europe_red <- data_europe[data_europe$summary != "Lev+Blade",]

```

We now check for homogeneity of variance in each group. We choose the
Brown-Forsythe test [@Brown1974] because the data have high numbers of
outliers. 

```{r bf-test1}

res.bf <- bf.test(length.width.thickness.2 ~ summary, data = data_europe_red)
summary(res.bf)

```

The p-value is under 0.003, we thus reject the hypothesis of homogeneity
of variances. Since our normal distributions do not have the same scale,
we use the Kruskall-Wallis test [@Kruskall1952] to see if at least one
group is significantly different from the others from the point of view
of their EFS ratio. 

```{r kw1}
res.kw <- kruskal.test(length.width.thickness.2 ~ summary, data = data_europe_red)
print(res.kw)
```

The p-value is well under 0.003, we can thus reject the hypothesis that
all distributions come from the same initial distribution. There is at
least one group that is significantly different from the rest. 

```{r plot-kw, fig.height=8}

par(mfrow = c(1,1), mar = c(10, 4.1, 4.1, 2.1) )
boxplot(length.width.thickness.2 ~ summary, data = data_europe_red, col = colors, las = 2,
        xlab = "", ylab = "sharp edge/flake size ratio")
text(x = .85 * par('usr')[2], y = .9 * par('usr')[4], paste0("Kruskal-Wallis p-value: ", res.kw$p.value), cex = 0.9)

```

### Data grouped by technocomplex

We now explore morphometric coherence of techno-complex categories. 

Before running statistical differentiation tests, we check for each
groups' distribution normality via a Lilliefors test. 

```{r normal2, fig.height=10}

par(mfrow = c(3,3))

for(i in 1:length(unique(data_europe$TCL))) {
  groupn <- data_europe[data_europe$TCL == unique(data_europe$TCL)[i], "length.width.thickness.2"]
  
  title <- paste(unique(data_europe$TCL)[i])
  
  #Shapiro Wilks test
  pvalue <- signif(lillie.test(groupn)$p.value, digits = 3)
  sample.size <- length(groupn)
  
  hist(groupn, breaks = seq(0,500,10), main = title, xlab = "sharp edge/flake size", col = cividis(3)[2])
  legend(x = .25 * par('usr')[2], y = .9 * par('usr')[4], 
        legend = c(paste0("Lilliefors test p-value: ", pvalue), paste0("sample size: ", sample.size)), bty = "n")
}

```

All groups present p-values under 0.003 [@Courtenay2024]. 

We now check for homogeneity of variance in each group. We choose the
Brown Forsythe test because the data have high numbers of outliers. 

```{r bf-test2}

res.bf <- bf.test(length.width.thickness.2 ~ TCL, data = data_europe_red)
summary(res.bf)

```

The p-value is under 0.003, we thus reject the hypothesis of homogeneity
of variances. Since our normal distributions do not have the same scale,
we use the Kruskall-Wallis test to see if at least one group is
significantly different from the others from the point of view of their
EFS ratio. 

```{r kw2}
res.kw <- kruskal.test(length.width.thickness.2 ~ TCL, data = data_europe_red)
print(res.kw)
```

The p-value is well under 0.003, we can thus reject the hypothesis that
all distributions come from the same initial distribution. There is at
least one group that is significantly different from the rest. 

```{r plot-kw2, fig.height=8}

order.tcl <- c("Quina Mousterian", "Mousterian", "MTA", "Chatelperronian", "Aurignacian", "Gravettian", "Solutrean", "Magdalenian")
data_europe$TCL <- factor(data_europe$TCL, levels = order.tcl)

par(mfrow = c(1,1), mar = c(10, 4.1, 4.1, 2.1) )
boxplot(length.width.thickness.2 ~ TCL, data = data_europe, col = cividis(3)[2], las = 2,
        xlab = "", ylab = "sharp edge/flake size ratio")
text(x = .85 * par('usr')[2], y = .9 * par('usr')[4], paste0("Kruskal-Wallis p-value: ", res.kw$p.value), cex = 0.9)

```

# References cited 